---
title: "Assignment 4 - Tidy Data"
author: "Mike Marin"
output: 
  html_document:
    toc: true
    toc_float: true
    #number_sections: true
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=7, fig.height=5)
```

```{r, warning = FALSE, message = FALSE, echo=FALSE}
library(gapminder)
library(tidyverse)
library(knitr)
library(DT)
```



# Introduction

This document is for Assignment #4.  The goal of this assignment is to work with __tidyr__ to make _tidy data_.  

The assignment consists of choosing an _option_ of different tasks to preform, and 3 different tasks.  These are done below.

Through peer-reviews, I've learned about making the table of contents _floating_, so I'm going to try that out here...I will note that it actually doesn't look very aesthetically pleasing, and some of the text is off screen and you have to scroll right to see it...but im keeping it this way for now, so that I have documented for myself an example of how to get this sort of TOC done.

# Exercise 1 - Univariate Data Reshaping

Here, we are given two options to choose from.  I am choosing to do __Univariate Option 1__.  I will carry out 3 tasks here.

## Part 1.1 - Widening

Here, we will work with the _gapminder_ dataset.  I will take the data (which is already in tidy format), and reshape it to have one row per year, and columns for life expectancy, for _Canada_, _America_, and _Mexico_.

This will be made into a tibble and displayed as a _datatable_.

```{r}
CMA_lifeExp <- gapminder %>% 
  filter(country=="Canada" 
         | country=="United States" 
         | country=="Mexico") %>% 
  pivot_wider(id_cols = year, names_from = country, 
              values_from = lifeExp) 
DT::datatable(CMA_lifeExp)
```


## Part 1.2 - Make a Plot

Here, we will use this _widened_ data to make a plot comparing the life expectancies of Canada, USA, and Mexico.

Since I have selected 3 countries, and I am interpreting the task as to plot x=country1 and y=country2, I will select one pair of countries (Canada and Mexico) for which to make the plot. (NOTE: the question could also be interpreted as wanting to plot life expectancies, over time, for multiple countries on the same plot.  I will _not_ be making such a plot)

```{r}
CMA_lifeExp %>% 
  ggplot(aes(x=Canada, y=Mexico)) +
  geom_point(size=3) +
  geom_line(size=2, alpha=0.2, colour="blue") +
  ggtitle("Canada vs Mexico Life Expectancy, from 1952-2007") +
  xlab("Canada Life Expectancy (years)") +
  ylab("Mexico Life Expectancy (years)" ) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggthemes::theme_economist()

```

As one would expect, there are highly correlated. 


## Part 1.3 - Re-Lengthen

Now, let's put the data back the way it was... in other words, let's re-lengthen it...

```{r Q1.3}
CMA_lifeExp %>% 
  pivot_longer(cols = c("Canada", 
                        "United States", "Mexico"), 
               names_to = "country", 
               values_to = "lifeExp") %>% 
  arrange(country) %>%
  DT::datatable()
```


# Exercise 2 - Multivariate Data Reshaping

Here, we are given two options to choose from.  I am choosing to do __Multivariate Option 1__.  I will carry out 3 tasks here.

I have chosen this task as it is a continuation of what was done in Exercise 1.


## Part 2.1 - Widening

Here, we will work with the _gapminder_ dataset.  I will make a tibble with one row per year, and columns for life expectancy and GDP per capita.

Like before, I will use countries of Canada, USA, and Mexico.  I will select numeric variables of _population_ and _gdpPerCapita_.


```{r}
CMA_lifeExp_multi <- gapminder %>% 
  filter(country=="Canada" 
         | country=="United States" 
         | country=="Mexico") %>% 
  pivot_wider(id_cols = year, names_from = country, 
              values_from = c("pop", "gdpPercap"))

DT::datatable(CMA_lifeExp_multi)
```


## Part 2.2 - Lengthening

Now, we will re-lengthen the data

```{r}
CMA_lifeExp_multi %>% 
  pivot_longer(cols = -year,
               names_to = c(".value", "country"),
               names_sep = "_" ) %>% 
  arrange(country) %>% 
  DT::datatable()
```



# Exercise 3 - Joining Tables

Here there are no choices to be made, we will work through all activities listed.  The goal is to work with _tidyr_ and _dplyr_ to work with _table joins_.

To do so, we will be working with the following made up data:

```{r, message=FALSE}
guest_data <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/attend.csv")
email_data <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/emails.csv")
```


First, let's just take a look at these data in tibbles, so we can see what we're working with, and how we are going to go about thigns...

```{r}
guest_data %>%
DT::datatable()
```


```{r}
DT::datatable(email_data)
```


We can see that the _email_ tibble has all guests that are part of the same party in one row, separated by commas, so first we should get each of them into their own row. We shoudl also rename the varaible __"guest"__ to __name__ so that the two tibbles have corresponding variable names. 

```{r separate email rows}
email_new <- email_data %>% 
  separate_rows(guest, sep = ", ") %>%
  rename(name=guest)
DT::datatable(email_new)
```

Ok, this should be easier to work with...

Also, worth noting that a _stupid_ __*white space*__ gave me so many head aches...the sep needed is ", ", not ",", as there is a space after the comma...jut a note for future to be midnful of this...


## Part 3.1 

The task here is:  For each guest in the guestlist (guest tibble), add a column for email address, which can be found in the email tibble.


```{r}
left_join(guest_data, email_new, by="name") %>%
  select(party, name, email) %>%
  DT::datatable()
```


We can see that some guests have not provided us with their email address...probably a good idea, so we don't spam them ;)


## Part 3.2 

Who do we have emails for, yet are not on the guest list.  And more importantly, why do we have their emails if they werent invited!  Wedding crashers!

```{r}
email_new %>%
  anti_join(guest_data, by="name") %>%
  DT::datatable()
```

Here we see our 3 wedding crashers: the Jones' and the Marshalls!


## Part 3.3

Here, we will make a guestlist that includes everyone we have emails for (in addition to those on the original guestlist).  That is, we will take the original guest list __(guest_data)__ , and add to it anything from the __(email_new)__ list, using a _fulljoin_.

```{r}
full_join(guest_data, email_new, by="name") %>%
  DT::datatable()
```





