---
title: "Assignment 4 - Tidy Data"
author: "Mike Marin"
output: 
  html_document:
    keep_md: true
    toc: true
    toc_float: true
    #number_sections: true
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=7, fig.height=5)
```

```{r, warning = FALSE, message = FALSE, echo=FALSE}
library(gapminder)
library(tidyverse)
library(knitr)
library(DT)
```



# Introduction

This document is for Assignment #4.  The goal of this assignment is to work with __tidyr__ to make _tidy data_.  

The assignment consists of choosing an _option_ of different tasks to preform, and 3 different tasks.  These are done below.

Through peer- reviews, I've learned about making the table of contents _foalting_, so I'm excited to try that out here...

# Exercise 1 - Univariate Data Reshaping

Here, we are given two options to choose from.  I am choosing to do __Univariate Option 1__.  I will carry out 3 tasks here.

## Part 1.1 - Widening

Here, we will work with the _gapminder_ dataset.  I will take the data (which is already in tidy format), and reshape it to have one row per year, and columns for life expectancy, for _Canada_, _America_, and _Mexico_.

This will be made into a tibble and displayed as a _datatable_.

```{r}
CMA_lifeExp <- gapminder %>% 
  filter(country=="Canada" 
         | country=="United States" 
         | country=="Mexico") %>% 
  pivot_wider(id_cols = year, names_from = country, 
              values_from = lifeExp) 
DT::datatable(CMA_lifeExp)
```


## Part 1.2 - Make a Plot

Here, we will use this _widened_ data to make a plot comparing the life expectancies of Canada, USA, and Mexico.

Since I have selected 3 countries, and I am interpreting the task as to plot x=country1 and y=country2, I will select one pair of countries (Canada and Mexico) for which to make the plot. (NOTE: the question could also be interpreted as wanting to plot life expectancies, over time, for multiple countries on the same plot.  I will _not_ be making such a plot)

```{r}
CMA_lifeExp %>% 
  ggplot(aes(x=Canada, y=Mexico)) +
  geom_point(size=3) +
  geom_line(size=2, alpha=0.2, colour="blue") +
  ggtitle("Canada vs Mexico Life Expectancy, from 1952-2007") +
  xlab("Canada Life Expectancy (years)") +
  ylab("Mexico Life Expectancy (years)" ) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggthemes::theme_economist()

```

As one would expect, there are highly correlated. 


## Part 1.3 - Re-Lengthen

Now, let's put the data back the way it was... in other words, let's re-lengthen it...

```{r Q1.3}
CMA_lifeExp %>% 
  pivot_longer(cols = c("Canada", 
                        "United States", "Mexico"), 
               names_to = "country", 
               values_to = "lifeExp") %>% 
  arrange(country) %>%
  DT::datatable()
```


# Exercise 2 - Multivariate Data Reshaping

Here, we are given two options to choose from.  I am choosing to do __Multivariate Option 1__.  I will carry out 3 tasks here.

I have chosen this task as it is a continuation of what was done in Exercise 1.


## Part 2.1 - Widening

Here, we will work with the _gapminder_ dataset.  I will make a tibble with one row per year, and columns for life expectancy and GDP per capita.

Like before, I willuse countries of Canada, USA, and Mexico.  I will select numeric variables of _population_ and _gdpPerCapita_.


```{r}
CMA_lifeExp_multi <- gapminder %>% 
  filter(country=="Canada" 
         | country=="United States" 
         | country=="Mexico") %>% 
  pivot_wider(id_cols = year, names_from = country, 
              values_from = c("pop", "gdpPercap"))

DT::datatable(CMA_lifeExp_multi)
```


## Part 2.2 - Lengthening

Now, we will relengthen the data

```{r}
CMA_lifeExp_multi %>% 
  pivot_longer(cols = -year,
               names_to = c(".value", "country"),
               names_sep = "_" ) %>% 
  arrange(country) %>% 
  DT::datatable()
```



# Exercise 3 - Joining Tables

Here there is no choice, we will work through all activities listed.  The goal is to work with _tidyr_ and _dplyr_ to work with _table joins_.

To do so, we will be working with the following made up data:

```{r, message=FALSE}
guest <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/attend.csv")
email <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/emails.csv")
```



## Part 3.1 

The task here is:  For each guest in the guestlist (guest tibble), add a column for email address, which can be found in the email tibble.






