---
title: "Untitled"
author: "Mike Marin"
date: '2019-09-17'
output: 
  beamer_presentation:
classoption: "aspectratio=169"
---


```{r LoadingLibraries, warning=FALSE, echo=FALSE}
library(tibble)
library(gapminder)
library(DT)
```

# Gapminder Data Summary 

## Data Description

The gapminder data contains information recorded for a number of countries

Variables recorded include: __Country__, __Continent__, __Year__, __Life Expectancy__, __Population__, __GDP Per Capita__

##

This is another slide added, without a title :)

I wanted to make sure I was able to add a slide, without giving it a title


# Some Univariate Summaries

## Univariate Summaries

It is of interest to take a look at each of the variables, and provide a quick summary, both numerically as well as graphically

## Summary Statistics


```{r, echo=FALSE}
summary(gapminder)
```


## Countires and Continents

Which countries are represented in the data?  

- In total there are 142 countries represented. They are not all listed here for the sake of space.  

- There are a total of 5 geopolitical continents recorded:
    + Africa
    + Asia
    +  Europe
    + Americas (North America and South America combined)
    + Oceania (Australia, New Zealand, and surrounding island nations)
    + Antarctica is not included in the data


## Years Recorded

The data was collected for a number of years.  The years the data was collected for were:



```{r, echo=FALSE}
levels(as.factor(gapminder$year))[1:7]
levels(as.factor(gapminder$year))[8:12]
```



<font size="3">
* Just adding some smaller font here for myself...
</font>

## Life Expectancy

* Making a figure smaller (should probably center it too)

```{r, warning=FALSE, echo=FALSE, fig.width=4, fig.height=3}
boxplot(gapminder$lifeExp, ylab="LifeExp", las=1)
summary(gapminder$lifeExp)
```


## GDP Per Capita


```{r, warning=FALSE, echo=FALSE, fig.width=3, fig.height=3, fig.align="center"}
boxplot(gapminder$gdpPercap, ylab="GDP/cap", las=1)
summary(gapminder$gdpPercap)
```

## log-GDP (because it's a Rate)

A quick note that since GDP/cap is a _**rate**_, it may be better to be exploring it on the log(ln)-scale

```{r, warning=FALSE, echo=FALSE, fig.width=3.5, fig.height=3.5, fig.align="center"}
boxplot(log(gapminder$gdpPercap), ylab="log-GDP/cap", las=1)
```


## Population

Here are words...

```{r, warning=FALSE, echo=FALSE, fig.width=3.5, fig.height=3.5, fig.align="center"}
boxplot(gapminder$pop, ylab="Population", las=1)
```


## log-Population

Again, it may be better to examine this on the log-scale, as populations tend to grow exponentially

```{r, warning=FALSE, echo=FALSE, fig.width=3.5, fig.height=3.5, fig.align="center"}
boxplot(log(gapminder$pop), ylab="Population", las=1)
```



# Some Bivariate Summaries

## Bivariate Summaries

We may be interested in examining relationships between some of the variables.  Following are a few of interest.

Note, we are ignoring the fact that the same countires have measurements taken for multiple years, for the time being...

## Life Expectancy and GDP

It is reasonable to hypothesize that these would be related.  Following is a visual examination of their relationship

```{r, echo=FALSE, warning=FALSE, fig.width=3.5, fig.height=3.5, fig.align="center"}
plot(gapminder$gdpPercap, gapminder$lifeExp)
```

## Pearson's correlation

While the relationship clearly isn't linear, let's calculate this anyway...

```{r, warning=FALSE}
cor(gapminder$gdpPercap, gapminder$lifeExp)
```

## Spearman's Correlation

Let's calculate Spearman's correlation to address the non-linearity, as it does appear to be a monotonic association:

```{r, warning=FALSE}
cor(gapminder$gdpPercap, gapminder$lifeExp, method="spearman")
```

This makes more sense, but still, we can try other things...

## Using log-GDP


Let's look at the same plot, but this time using the log-GDP-per-capita, as it makes sense to examine it on this scale

```{r, echo=FALSE, warning=FALSE, fig.width=3.5, fig.height=3.5, fig.align="center"}
plot(log(gapminder$gdpPercap), gapminder$lifeExp)
```

## That Looked Like a Fix!

That looks like it would be easier to model.  Let's also take a quick look at Pearson's correlation when we use log-GDP...

```{r, warning=FALSE, fig.width=3.5, fig.height=3.5, fig.align="center"}
cor(log(gapminder$gdpPercap), gapminder$lifeExp, method="pearson")
```

# The End?

## Yes, The End!

This will be the ending here

* There wasn't much focus on the content of presentation

- I wasn't consistent with using "echo=F" or not

    - But that's because I wanted to focus on learning more about how it formats slides, etc

- Some basic things were tried

- I don't really care about my grade, I just want to learn about the course material




