---
title: "Untitled"
author: "Mike Marin"
date: '2019-09-17'
output: pdf_document
always_allow_html: yes
---
# The Gapminder Data



```{r LoadingLibraries, warning=FALSE, echo=FALSE}
library(tibble)
library(gapminder)
library(DT)
```

## Variables Collected

Here, we can see the __variables__ collected in the data set:

```{r, echo=FALSE}
names(gapminder)
```

## Some Basic Summaries


The data dimensions are as folows:

```{r}
dim(gapminder)
```

## The Full Data

To explore the full data, you can play around here:

```{r}
# the datatable makes it easier to do something in HTML
datatable(as_tibble(gapminder))
```

# Some Univariate Summaries

It is of interest to take a look at each of the variables, and provide a quick summary, both numerically as well as graphically

## A Quick Glance

Following are basic univariate summaries for each of the variables:

```{r, echo=FALSE}
summary(gapminder)
```

## Countires and Continents

Which countries are represented in the data?  

- In total there are 142 countries represented. They are not all listed here for the sake of space.  

- There are a total of 5 geopolitical continents recorded:
    + Africa
    + Asia
    +  Europe
    + Americas (made up of North America and South America are combined)
    + Oceania (Australia, New Zealand, and surrounding island countries)
    + Antarctica is not included in the data

## Years Recorded

The data was collected for a number of years.  The years the data was collected for were:

```{r, echo=FALSE}
levels(as.factor(gapminder$year))
```


## Life Expectancy, GDPperCap, and Population

These are all numeric variables, and all likely related to one another.  Following are basic summaries and plots for each:

### Life Expectancy

```{r, warning=FALSE}
boxplot(gapminder$lifeExp, ylab="LifeExp", las=1)
summary(gapminder$lifeExp)
```

Wow, 23 years life expectancy?!  Where did that come from?

```{r, echo=FALSE}
gapminder[gapminder$lifeExp <24, ]
```

Oh, ok, we all know that there was a  *__LOT__* going on there at that time!

### GDP Per Capita


```{r, warning=FALSE}
boxplot(gapminder$gdpPercap, ylab="GDP/cap", las=1)
summary(gapminder$gdpPercap)
```

And a quick note that since GDP/cap is a rate, it may be better to be exploring it on the log(ln)-scale

```{r, warning=FALSE}
boxplot(log(gapminder$gdpPercap), ylab="log-GDP/cap", las=1)
summary(log(gapminder$gdpPercap))
```

### Population

```{r, warning=FALSE}
boxplot(gapminder$pop, ylab="Population", las=1)
summary(gapminder$pop)
```

Again, it may be better to examine this on the log-scale, as populations tend to grow exponentially

```{r, warning=FALSE}
boxplot(log(gapminder$pop), ylab="Population", las=1)
summary(log(gapminder$pop))
```




# Some Bivariate Summaries

We may be interested in examining relationships between some of the variables.  Following are a few of interest.

Note, we are ignoring the fact that the same countires have measurements taken for multiple years, for the time being...

## Life Expectancy and GDP

It is reasonable to hypothesize that these would be related.  Following is a visual examination of their relationship

```{r, echo=FALSE, warning=FALSE}
plot(gapminder$gdpPercap, gapminder$lifeExp)
```

Pearson's correlation:

```{r, warning=FALSE}
cor(gapminder$gdpPercap, gapminder$lifeExp)
```

It's not surprising to see they are associated.  It also isn't surprising to see that it is a non-liear relationship.

First, let's calculate Spearman's correlation to address the non-linearity, as it does appear a monotonic association:

```{r, warning=FALSE}
cor(gapminder$gdpPercap, gapminder$lifeExp, method="spearman")
```

yup, it looks much more reasonable


Let's look at the same plot, but this time using the log-GDP-per-capita, as it makes sense to examine it on this scale

```{r, echo=FALSE, warning=FALSE}
plot(log(gapminder$gdpPercap), gapminder$lifeExp)
```

That looks like it would be easier to model.  Let's also take a quick look at Pearson's correlation when we use log-GDP...

```{r, warning=FALSE}
cor(log(gapminder$gdpPercap), gapminder$lifeExp, method="pearson")
```


Let's not get carried away...we will stop here

## Continent and GDP

It is also reasonable to explore if GDP per captia varies by continent (again, ignoring the fact that we have measurements for multiple years)

Let's look at a plot of this

```{r, echo=FALSE, warning=FALSE}
boxplot(gapminder$gdpPercap~gapminder$continent, ylab="GDP/cap", las=1)
```

We can see an associaition.  And as before, it may be more reasonable to examine GDP per capita on the _log-scale_ as this variable is a __rate__

Let's look at that here...

```{r, echo=FALSE, warning=FALSE}
boxplot(log(gapminder$gdpPercap)~gapminder$continent, ylab="log-GDP/cap", las=1)
```


## No More...

I wont get carried away, as the point of this exercise is working with R Markdown and Git(Hub), and not the actual data analysis...although, this is a *__VERY__* interesting dataset, and I will be exploring it further on my own, for myself :)